Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-22 manuscript/response_to_reviews.Rmd

*Current text, with phrase of concern in bold: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]." *

**To address the question about species, we have added a new figure (*Fig. 3*) showing how $Rt$ varied across species in each of the three droughts.**

- Line 331-333: not so sure about this. It can also mean that there is no causal relationship between Rt and covariates, mostly because radial growth Rt should not be used as an integral proxy of tree resistance to drought, such as it is often used now in the literature.

**Our analysis of $Rt_{ARIMA}$ (described above) confirms the findings based on analysis of $Rt$.**

@@

L160: you mean a log-log regression, right?

**Yes. fixed.**

L181: N=3 individuals per species is really quite low. In addition, by sampling up to 8 m height you would probably be sampling less exposed branches from taller trees, which could confound your results.

**There were significant differences among species for all sampled traits, and intraspecific SD was generally low with respect to mean values and interspecific variation. We have added a figure to the SI (Fig. S4) showing variation in traits across species. We also note that a sample size of n=3 is not unprecendented for this type of analysis [@medeiros_extensive_2019]. More samples would be ideal, but was not feasible given the time and resources available. **

@@

L223: clarify that the -4.53 value corresponds to July.

**This sentence has been re-written: "The 1999 drought was preceded by wetter than average conditions until the previous June, but $PDSI$ plummeted below -3.0 in October 1998 and remained below this threshold through August 1999."**

L228-234: please clarify model structure. I understand you included tree as a random factor only when the three droughts were analyzed together, right? It would be nice to know what was the contribution of random factors to model fit (only one R2 is shown in Table 5, and it is unclear whether it corresponds to marginal or conditional values).

**We have substantially revised this section and expect that it should be far more clear. The text in question now reads:**

**"The general statistical model for hypothesis testing was a mixed effects model, implemented in the lme4 package in R [@R-lme4]. In the multi-year model, we included a random effect of tree nested within species and a fixed effect of drought year to represent the combined effects of differences in drought characteristics.  Individual year models included a random effect of species.  All models included fixed effects of independent variables of interest (Tables 1,3) as specified below."**

**We have specified that R2 corresponds to conditional values (footnotes of former table 5/ current table S6, and table S7).**

@@

L238, 242, 244: it is unclear to me why in some cases you use a dAICc of 1 and sometimes of 2.

**dAICc = 2 was consistently used to test for significance, whereas dAICc=1 was used as a preliminary threshhold for variable selection. We now clearly specify this in the footnote of table S4 (former table 4) and in the text: "Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table S4). We note that we did not use the $\Delta$AICc $\ge$ 1.0 criterion as a test of significance, but rather of whether the variable had enough influence to be considered as a *candidate* variable in full models."**

@@

L287-299: it would be very useful to use scatterplots to show the effects of traits on Rt in a figure.

**We have added the specified figure (Fig. S6). In addition, we have added 3 related figures illustrate how traits and $Rt$ vary across species (Figs. S4 and 3, respectively) and how $Rt$ responds to  $PLA_{dry}$ and $\pi_{tlp}$.**

@@

L297: TLP was significant for 1966 according to Table 4.

**This text was removed.**

L301-302: I would tone down this, as your models explained a rather low % of the variation in Rt.

**Tree-ring data are inherently noisy, particularly in more mesic forests. In large part, this is due to strong neighborhood effects, along with variable allocation to ring growth around the circumference of the tree. These factors contribute to the low R^2^, but do not negate the significance of our findings.**

@@

L305-306: not for the 1977 and 1999 droughts, according to Table 5.

**This text was removed.**

L314-316: or simply that your sample size was rather low. See also main comments above.

**Our sample size of tree-ring data is quite good, and we believe the traits sample size to be sufficient (see above). We have now tried numerous iterations of this analysis, and the result described here holds. Biologically, it also makes sense that the importance of height, TWI, and traits would vary somewhat across droughts, given that every drought is different.**

### 2020-07-22 manuscript/climate_sensitivity_SI.Rmd
"leaf hydraulic traits" --> "leaf drought tolerance traits"
I think I got everything. Closes #126
finished first pass at reviewer responses!!!!

### Supplementary Information
## Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

@@

>>>>

### Supplementary Information
## Tree height and leaf drought tolerance traits shape growth responses across droughts in a temperate broadleaf forest

@@

![**Figure S6. Visualization of best model, with data, for all droughts combined.** Model coefficients are given in Table S6.](tables_figures/publication/FigureS6_best_model_with_data.png)

### 2020-07-22 manuscript/climate_sensitivity_draft.Rmd
finished first pass at reviewer responses!!!!

- Individual-level drought resistance decreased with tree height, which was strongly correlated with *crown exposure*. The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index. Resistance was greater among species whose leaves experienced less shrinkage upon desiccation and lost turgor (wilted) at more negative water potentials.

@@

Analyses focused on testing the predictions presented in Table 1 with $Rt$ as the response variable, and then repeated using $Rt_{ARIMA}$ as the response variable. 
Models were run for all drought years combined and for each drought year individually. 
The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [@R-lme4], using tree nested within species as a random effect, and independent variables as specified below. 
In the multi-year model, we also included a fixed effect of drought year to represent the combined effects of differences in drought characteristics.
We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
AICc refers to a corrected version of AICc, and is best suited for small data sizes [see @brewer_relative_2016].

@@

Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Fig. **4**). 
Specifically, $ln[H]$ appeared, with a negative coefficient, in the best model (($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S6-S7). 
The same held true for 1966 individually.
For the 1977 drought, $ln[H]$ did not appear in the best model, but was included, with a negative coefficient, among the top models--*i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (Tables 1, S6-S7). 
For the 1999 drought, $ln[H]$ had no significant effect.

$Rt$ had a significantly negative response to $ln[TWI]$ in all drought years combined and in 1977 and 1999 individually (Fig. **4**, Table S6). 
When $Rt_{ARIMA}$ was used as the response variable, the effect was significant in all drought years combined and in 1977, and a negative effect of $ln[TWI]$ was included in some of the models in 1966 and 1999 (Table S7).
This negates the idea that trees in moist microsites would be less affected by drought. 
Nevertheless, we tested for a $ln[H] *ln[TWI]$ interaction, a negative sign of which could indicate that smaller trees (presumably with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table. 
This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top models in which it appeared (Tables 1, S6-S7).
This term did appear, however, with a positive coefficient in the best $Rt_{ARIMA}$ model for all years combined (Table S7).

@@

Species traits... (Table 2, Fig. **S4**)
Responses varied across species and by drought (Fig. **3**). 
Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).

@@

In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were both negatively correlated to drought resistance (Fig. **4**; Tables 1,S4-S7). 
Both had consistent signs across all droughts, and their inclusion at least marginally improved the model ($\Delta$AICc > 1.0) for at least one of the three droughts (Table S4), qualifying them as candidate variables for the full model. 
$PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S6-S7). 
For 1977 and 1999, it was included with a negative coefficient in some of the top models (Tables S6-S7).
$\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig. **4**; Table S6). 
It was also included in some of the top models for 1999 (Tables S6-S7).

@@

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig. **1a**, @helcoski_growing_2019) and had substantial but not devastating impacts on tree growth (Figs. **1b**). These droughts were classified as severe ($PDSI$ < -3.0; 1977) or extreme ($PDSI$ < -4.0; 1966, 1999) at our site and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]. However, extreme, multiannual droughts such as the so-called “megadroughts” of this type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016]. Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. **S1**), severely stressed a larger portion of trees (Fig. **1b**). The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. **1b**), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]. It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-25% for each individual drought (Table S6).

>>>>

- Individual-level drought resistance decreased with tree height, which was strongly correlated with exposure to higher evaporative demand and solar radiation. The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index. Resistance was greater among species whose leaves experienced less shrinkage upon desiccation and lost turgor (wilted) at more negative water potentials.

@@

Analyses focused on testing the predictions presented in Table 1 with $Rt$ as the response variable, and then repeated using $Rt_{ARIMA}$ as the response variable. 
Models were run for all drought years combined and for each drought year individually. 
The general statistical model for hypothesis testing was a mixed effects model, implemented in the lme4 package in R [@R-lme4].
In the multi-year model, we included a random effect of tree nested within species and a fixed effect of drought year to represent the combined effects of differences in drought characteristics.
Individual year models included a random effect of species.  
All models included fixed effects of independent variables of interest (Tables 1,3) as specified below. 
We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
AICc refers to a corrected version of AICc, and is best suited for small data sizes [see @brewer_relative_2016].

@@

Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Figs. **4, S6**). 
Specifically, $ln[H]$ appeared, with a negative coefficient, in the best model (($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S6-S7). 
The same held true for 1966 individually.
For the 1977 drought, $ln[H]$ did not appear in the best model, but was included, with a negative coefficient, among the top models--*i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (Tables 1, S6-S7). 
For the 1999 drought, $ln[H]$ had no significant effect.

$Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined and in 1977 and 1999 individually (Figs. **4, S6**, Table S6-S7). 
The effect was also significant for 1977 and 1999 individually (Fig. **4**, Table S6).
When $Rt_{ARIMA}$ was used as the response variable, the effect was significant in 1977, and included in some of the top models in 1966 and 1999 (Table S7).
This negates the idea that trees in moist microsites would be less affected by drought. 
Nevertheless, we tested for a $ln[H] *ln[TWI]$ interaction, a negative sign of which could indicate that smaller trees (presumably with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table. 
This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top models in which it appeared (Tables 1, S6-S7).
This term did appear, however, with a positive coefficient in the best $Rt_{ARIMA}$ model for all years combined (Table S7).

@@

Species traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$) varied significantly among species (all p<0.05 in ANOVA; Table 2, Fig. **S4**).
Drought resistance also varied across species and by drought (Fig. **3**). 
Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).

@@

In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were both negatively correlated to drought resistance (Figs. **4, S6**; Tables 1,S4-S7). 
Both had consistent signs across all droughts, and their inclusion at least marginally improved the model ($\Delta$AICc > 1.0) for at least one of the three droughts (Table S4), qualifying them as candidate variables for the full model. 
$PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S6-S7). 
For 1977 and 1999, it was included with a negative coefficient in some of the top models (Tables S6-S7).
$\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig. **4**; Table S6). 
It was also included in some of the top models for 1999 (Tables S6-S7).

@@

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig. **1a**, @helcoski_growing_2019) and had substantial but not devastating impacts on tree growth (Figs. **1b**). These droughts were classified as severe ($PDSI$ < -3.0; 1977) or extreme ($PDSI$ < -4.0; 1966, 1999) at our site and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]. However, extreme, multiannual droughts such as the so-called “megadroughts” of this type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016]. Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. **S1**), severely stressed a larger portion of trees (Fig. **1b**). The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. **1b**), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]. It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-25% for each individual drought (Fig. **S6**; Table S6).
